Individual and Differential Harm in Redistricting

### 2022-05-24 paper/paper.Rmd
work on citations

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [cite], is perhaps the most consistently visible manifestation of this tendency.
The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@katz2020, need more].
In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [cite].
The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [cites].
At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [cite].
Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see \_\_\_\_].

@@

This framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidates [@2004vieth; @2018league; @2019lewis].
A more detailed discussion of the connections between the proposed framework existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.

@@

[^8]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
    One of the authors (Kenny) provided research support to an expert witness for the plaintiffs.

@@

The table below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for \_\_ states which cover a wide range of partisan leans and number of seats.
The simulations come from [alarm cite].

>>>>

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp. 72--73], is perhaps the most consistently visible manifestation of this tendency.
The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
Other types of gerrymandering exist as well.
In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process].
Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see @siegel2013educational; @vargas2021racial].

@@

This framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
A more detailed discussion of the connections between the proposed framework existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.

@@

[^8]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.

@@

The table below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for \_\_ states which cover a wide range of partisan leans and number of seats.
The simulations are provided by the 50-State Simulation Project [@mccartan2022fiftystates].
